---
output: github_document
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(cache = FALSE, autodep = TRUE,  warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, tidy = TRUE, fig.width = 9, fig.height = 6, purl=TRUE, fig.show = "hold", cache.lazy = FALSE, fig.path = "README_figs/README-")
```

# Introduction

Dependence between test statistic is known to render statistical inference more variable. This is caused by the random nature of the null distribution under dependence. The aim of this _reconsi_ pacakage is to provide Resampling RAndom Null distributions 
for Simultaneous Inference to estimate these random functions for improved inference. It can reduce the variability of the false discovery proportion, while increasing the power. Any type of test statistic can be supplied, only Wilcoxon rank sum test and two sample t-test are natively implemented. Both **permutations** and **bootstrapping** are implemented.

A short tutorial is given here, more detailed instructions can be found in the vignette.

# Installation

The package can be installed using the following commands:

```{r install, eval = FALSE}
library(devtools)
install_github("CenterForStatistics-Ugent/reconsi")
```

and loaded as

```{r loadRransi}
library(reconsi)
```

# General use

We illustrate the general use of the package on a synthetic dataset. The default 
Wilcoxon rank-sum test is used.

```{r syntData}
#Create some synthetic data with 90% true null hypotheses
 p = 200; n = 50
 x = rep(c(0,1), each = n/2)
 mat = cbind(
 matrix(rnorm(n*p/10, mean = 5+x),n,p/10),
 matrix(rnorm(n*p*9/10, mean = 5),n,p*9/10)
 )
 #Provide just the matrix and grouping factor, and test using the random null
 fdrRes = reconsi(mat, x)
 #The estimated tail-area false discovery rates.
 estFdr = fdrRes$Fdr
```

The method provides an estimate of the proportion of true null hypothesis, which 
is close to the true 90\%.

```{r p0}
fdrRes$p0
```

The result of the procedure can be represented graphically as follows:

```{r plotNull}
plotNull(fdrRes)
```

It is also possible to provide a custom test function, which must accept at least a 'y' response variable and a 'x' grouping factor. Additionally, a distribution function should be supplied for transformation through quantiles to z-values.

```{r customFunction}
 #With another type of test
fdrResLm = reconsi(mat, x, B = 5e1,
                      test = function(x, y){
                          fit = lm(y~x)
                          c(summary(fit)$coef["x","t value"], fit$df.residual)},
                      distFun = function(q){pt(q = q[1], df = q[2])})
```

## Case study

We illustrate the package using an application from microbiology. The species
composition of a community of microorganisms can be determined through 
sequencing. However, this only yields compositional information, and knowledge 
of the population size can be acquired by cell counting through flow cytometry.
Next, the obtained species compositions can multiplied by the total population 
size to yield approximate absolute cell counts per species. Evidently, this 
introduces strong correlation between the tests due to the common factor. In 
other words: random noise in the estimation of the total cell counts will affect 
all hypotheses. Therefore, we employ permutations to estimate an 
experiment-specific random null distribution, that will account for this 
dependence.

The dataset used is taken from Vandeputte _et al._, 2017 (see [manuscript](https://www.ncbi.nlm.nih.gov/pubmed/29143816)), a study on gut microbiome 
in healthy and Crohn's disease patients. The test looks for differences in absolute abundance between healthy and diseased patients. It relies on the [phyloseq](https://bioconductor.org/packages/release/bioc/html/phyloseq.html) package, which is the preferred way to interact with our machinery for microbiome data.

```{r Vandeputte}
testVanDePutte = testDAA(Vandeputte, "Health.status", "absCountFrozen", B = 1e2L)
```

The estimated tail-area false discovery rates can then simply be extracted as

```{r vandeputteFdr}
FdrVDP = testVanDePutte$Fdr
quantile(FdrVDP)
```

An approximation of the correlation matrix of the test statistics can be drawn as follows:

```{r approxCor}
plotCperm(testVanDePutte$statsPerm)
```

This is the correlation matrix of binned test statistics, whereby yellow indicates negative correlations between bin counts and blue positive correlations. Each pixel represents a combination of two bins. It is clear that counts in the left and right tail are negatively correlated, and bin counts close together are positively correlated. This is a consequence of the random null distribution shifting left and right.
